---
output: github_document
bibliography: vignettes/references.json
link-citations: true
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->
<!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->
<!-- Variables marked with double curly braces will be transformed beforehand: -->
<!-- `packagename` is extracted from the DESCRIPTION file -->
<!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = file.path("man", "figures", "README-"),
  out.width = "100%"
)
```

# _{{ packagename }}_: Methods for simulating and analysing the size and length of transmission chains from branching process models <img src="man/figures/logo.svg" align="right" height="130" />

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/epiverse-trace/epichains/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/epichains/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/github/epiverse-trace/epichains/branch/main/graphs/badge.svg)](https://codecov.io/github/epiverse-trace/epichains)
[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

_{{ packagename }}_ is an R package to simulate, analyse, and visualize the size 
and length of branching processes with a given offspring distribution. These 
models are often used in infectious disease epidemiology, where the chains represent chains of
transmission, and the offspring distribution represents the distribution of 
secondary infections caused by an infected individual. 

_{{ packagename }}_ re-implements [bpmodels](https://github.com/epiverse-trace/bpmodels/)
by providing bespoke functions and data structures that allow easy
manipulation and interoperability with other Epiverse-TRACE packages, for example, [superspreading](https://github.com/epiverse-trace/superspreading/) and [epiparameter](https://github.com/epiverse-trace/epiparameter/), and potentially some existing packages for handling transmission chains, for example, [epicontacts](https://github.com/reconhub/epicontacts).

_{{ packagename }}_ is developed at the [Centre for the Mathematical Modelling of Infectious Diseases](https://www.lshtm.ac.uk/research/centres/centre-mathematical-modelling-infectious-diseases) at the London School of Hygiene and Tropical Medicine as part of the [Epiverse Initiative](https://data.org/initiatives/epiverse/).

## Installation

The latest development version of the _{{ packagename }}_ package can be installed via

```{r include=TRUE,eval=FALSE}
# check whether {pak} is installed
if (!require("pak")) install.packages("pak")
pak::pak("{{ gh_repo }}")
```

To load the package, use

```{r eval=TRUE}
library("epichains")
```

## Quick start

_{{ packagename }}_ provides two main functions:

* `simulate_chains()`: simulates transmission chains using a simple
branching process model that accepts an index number of cases that seed
the outbreak, a distribution of offspring per case, and a chain statistic
to track (size or length/duration). It optionally accepts other population
related inputs such as the population size (defaults to Inf) and percentage
of the population initially immune (defaults to 0). This function returns
an object with columns that track information on who infected whom, the
generation of infection and, if a generation time function is specified, the
time of infection.

* `simulate_summary()`: provides a performant version of `simulate_chains()`
that only tracks and return a vector of realized chain sizes or
lengths/durations for each index case without details of the infection tree.

* `likelihood()`: calculates the loglikelihood (or likelihood, depending
on the value of `log`) of observing a vector of transmission chain sizes or
lengths.

The objects returned by the `simulate_*()` functions can be summarised with
`summary()` and aggregated into a `<data.frame>` of cases per time or generation
with `aggregate()`. Aggregated results can also be passed on to `plot()` with
its own arguments to customize the resulting plots. 

Each of the listed functionalities is demonstrated in detail
in the ["Getting Started" vignette](https://epiverse-trace.github.io/epichains/articles/epichains.html).

## Package vignettes

The theory behind the models provided here can be
found in the [theory vignette](https://epiverse-trace.github.io/epichains/articles/theoretical_background.html).

We have also collated a bibliography of branching process applications in 
epidemiology. These can be found in the [literature vignette](https://epiverse-trace.github.io/epichains/articles/branching_process_literature.html).

Specific use cases of _{{ packagename }}_ can be found in 
the [online documentation as package vignettes](https://epiverse-trace.github.io/epichains/), under "Articles".

## Reporting bugs 

To report a bug please open an [issue](https://github.com/epiverse-trace/epichains/issues/new/choose).

## Contribute

We welcome contributions to enhance the package's functionalities. If you 
wish to do so, please follow the [package contributing guide](https://github.com/epiverse-trace/epichains/blob/main/.github/CONTRIBUTING.md).

## Code of conduct

Please note that the _{{ packagename }}_ project is released with a [Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md). 
By contributing to this project, you agree to abide by its terms.

## Citing this package

```{r message=FALSE, warning=FALSE}
citation("epichains")
```
