---
title: "Getting started with epichains"
author: "James Azam"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting started with epichains}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

## Functionality

`epichains` currently has 4 core functions:

* `simulate_tree()`: simulate transmission trees from a given number of chains.
* `simulate_tree_from_pop()`: simulate transmission trees from a given number 
  population size and initial immunity.
* `simulate_summary()`: simulate a vector of observed transmission chains 
  sizes/lengths from a given number of chains.
* `likelihood()`: estimate the likelihood/loglikelihood of observing
  chains of given sizes/lengths.

### Object-orientation

#### Classes

* An `epichains` class:
  * superclass of `data.frame` with attributes for tracking `chain_type` as: 
    * `chains_tree`, if returned from `simulate_tree()` or 
    `simulate_tree_from_pop()`
    * `chains_vec`, if returned from `simulate_summary()`.
* An `epichains_aggregate_df` class:
  * superclass of `data.frame` with attributes for tracking if aggregation is 
  done over "time", "generation" or "both". Useful for `plot` method dispatch 
  (see methods section below).

#### Methods

* `print()`
* `summary()`
* `aggregate()`

## Demo

### Printing and summary
```{r include=TRUE,echo=TRUE}
library(epichains)
# Using `simulate_tree()`
tree_from_pois_offspring <- simulate_tree(
  nchains = 10,
  offspring_dist = "pois",
  serials_sampler = function(x) 3,
  lambda = 2,
  stat_max = 10
)

tree_from_pois_offspring # print the output

# Using simulate_summary()
summary_sim <- simulate_summary(
  nchains = 50, offspring_dist = "pois",
  statistic = "length", lambda = 2,
  stat_max = 10
)

summary_sim # print the output

# Using `simulate_tree_from_pop()`

# Simulate with poisson offspring
tree_from_pop_pois <- simulate_tree_from_pop(
  pop = 1000,
  offspring_dist = "pois",
  offspring_mean = 0.5,
  serial_sampler = function(x) 3
)

tree_from_pop_pois # print the output

# Simulate with negative binomial offspring
tree_from_pop_nbinom <- simulate_tree_from_pop(
  pop = 1000,
  offspring_dist = "nbinom",
  offspring_mean = 0.5,
  offspring_disp = 1.1,
  serial_sampler = function(x) 3
)

tree_from_pop_nbinom # print the output

# Likelihoods

chain_sizes <- c(1, 1, 4, 7)
likelihood(
  chains = chain_sizes, statistic = "size",
  offspring_dist = "pois", nsim_obs = 100,
  lambda = 0.5
)
```

### Aggregation
```{r include=TRUE,echo=TRUE}
# aggregate by time
aggregate(tree_from_pop_pois, "time")

# aggregate by generation
aggregate(tree_from_pop_pois, "generation")

# aggregate by both time and generation
aggregate(tree_from_pop_pois, "both")
```
