---
title: "Number of generations to achieve a set outbreak size"
author: "James Azam"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Number of generations to achieve a set outbreak size}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)

```

```{r packages}
library(epichains)
```

# Outline

* Question: At what generation does an outbreak reach a size of 100, given that
  it was started by a single individual with R0 = 1.5, and individual
  transmission, $k$, at levels of 0.01, 0.1, 0.5, 1, 4, and Infinity?
* Method: 
  * We use a negative binomial model with $mu = 1.5$ and
  $size = c(0.01, 0.1, 0.5, 1, 4, Inf)$.
  * We run 10, 000 simulations per each combination of $mu$ and $size$
  * We aggregate chain sizes by generation and extract the generation
    where cumulative chain size = 100.
  * We find the percentage of simulations that reached the 100 case size
    threshold.
* Outcome: The distribution (boxplot) of the realized generations per value of
  $k$. This reproduces Fig 2c of @lloyd-smith2005.

# References
