---
title: "Projecting infectious disease incidence: a COVID-19 example"
author: "James Azam, Sebastian Funk"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteIndexEntry{Projecting infectious disease incidence: a COVID-19 example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)

```

## Overview

Branching processes can be used to project stochastic infectious disease trends in time
provided we can characterize the distribution of times between
successive cases (serial interval), and the distribution of secondary cases
produced by a single individual (offspring distribution). Such simulations can
be achieved in _epichains_ with the `simulate_chains()` function and
@pearson2020, and @abbott2020 illustrate its application to COVID-19.

The purpose of this vignette is to use early data on COVID-19 in South Africa 
[@marivate2020] to illustrate how _epichains_ can be used to project an 
outbreak. 

Let's load the required packages

```{r packages, include=TRUE}
library("epichains")
library("dplyr")
library("ggplot2")
library("lubridate")
```

## Data

Included in _epichains_ is a cleaned time series of the first 15 days of 
the COVID-19 outbreak in South Africa. This can be loaded into 
memory as follows: 
```{r data}
data("covid19_sa", package = "epichains")
```

We will use the first $5$ observations for this demonstration. We will assume
that all the cases in that subset are imported and did not infect each other. 

Let us subset and view that aspect of the data.
```{r view_data}
seed_cases <- covid19_sa[1:5, ]
head(seed_cases)
```

## Setting up the inputs  

We will now proceed to set up `simulate_chains()` for the simulations.

### Onset times 

`simulate_chains()` requires a vector of seeding times, `t0`, for each
chain/individual/simulation.

To get this, we will use the observation date of the index case as the
reference and find the difference between the other observed dates and the reference. 
```{r linelist_gen, message=FALSE}
days_since_index <- as.integer(seed_cases$date - min(seed_cases$date))
days_since_index
```

Using the vector of start times from the time series, we will then 
create a corresponding seeding time for each individual, which we'll call
`tf`.
```{r t0_setup}
t0 <- rep(days_since_index, seed_cases$cases)
t0
```

### Generation time

In epidemiology, the generation time (also called the generation interval) is
the duration between successive infectious events in a chain of transmission.
Similarly, the serial interval is the duration between observed symptom onset
times between successive cases in a transmission chain. The generation
interval is often hard to observe because exact times of infection are hard to
measure hence, the serial interval is often used instead. Here, we
use the serial interval and interpret the simulated case data to represent
symptom onset.

In this example, we will assume based on COVID-19 literature that the 
serial interval, S, is log-normal distributed with parameters, 
$\mu = 4.7$ and $\sigma = 2.9$ [@pearson2020]. The log-normal distribution is
commonly used in epidemiology to characterise quantities such as the serial
interval because it has a large variance and can only be positive-valued
[@nishiura2007; @limpert2001].

Note that when the distribution is described this way, it means $\mu$ and
$\sigma$ are the expected value and standard deviation of the natural
logarithm of the serial interval. Hence, in order to sample the
"back-transformed" measured serial interval with expectation/mean, $E[S]$
and standard deviation, $SD [S]$, we can use the following parametrisation:

\begin{align}
E[S] &= \ln \left( \dfrac{\mu^2}{(\sqrt{\mu^2 + \sigma^2}} \right) \\

SD [S] &= \sqrt {\ln \left(1 + \dfrac{\sigma^2}{\mu^2} \right)}
 
\end{align}

See ["log-normal_distribution" on Wikipedia](https://en.wikipedia.org/wiki/Log-normal_distribution) for a
detailed explanation of this parametrisation.

We will now set up the generation time function with the appropriate inputs.
We adopt R's random lognormal distribution generator (`rlnorm()`) that
takes `meanlog` and `sdlog` as arguments, which we define with the
parametrisation above as `log_mean()` and `log_sd()` respectively and wrap it in 
the `generation_time_fn()` function. Moreover, `generation_time_fn()` takes one
argument `n` as is required by _epichains_ (See `?epichains::simulate_chains`),
which is further passed to `rlnorm()` as the 
first argument to determine the number of observations to sample
(See `?rlnorm`).
```{r generation_time_setup, message=FALSE}
mu <- 4.7
sgma <- 2.9

log_mean <- log((mu^2) / (sqrt(sgma^2 + mu^2)))  # log mean
log_sd <- sqrt(log(1 + (sgma / mu)^2)) # log sd

#' serial interval function
generation_time <- function(n) {
  gt <- rlnorm(n, meanlog = log_mean, sdlog = log_sd)
  return(gt)
}
```

### Offspring distribution

Let us now set up the offspring distribution, that is the distribution that
drives the mechanism behind how individual cases infect other cases. The
appropriate way to model the offspring distribution is to capture both the
population-level transmissibility ($R0$) and the individual-level heterogeneity
in transmission ("superspreading"). The negative binomial distribution is
commonly used in this case [@lloyd-smith2005].

For this example, we will assume that the offspring distribution is 
characterised by a negative binomial with $mu = 2.5$ [@abbott2020] and 
$size = 0.58$ [@wang2020]. 

```{r nbinom_args_setup, message=FALSE}
mu <- 2.5
size <- 0.58
```

In this parameterization, $mu$ 
represents the $R_0$, which is defined as the average number of 
cases produced by a single individual in an entirely susceptible population. 
The parameter $size$ represents superspreading, that is, the degree of 
heterogeneity in transmission by single individuals.

### Simulation controls

For this example, we will simulate outbreaks that end $21$ days after the last
date of observations in the `seed_cases` dataset.
```{r time_args_setup, message=FALSE}
#' Date to end simulation
projection_window <- 21
tf <- max(days_since_index) + projection_window
tf
```

`simulate_chains()` is stochastic, meaning the results are different every
time it is run for the same set of parameters. We will, therefore, run the
simulations $100$ times and aggregate the results.

Let us specify that.
```{r sim_reps_setup}
#' Number of simulations
sim_rep <- 100
```

Lastly, since, we have specified that $R0 > 1$, it means the epidemic could
potentially grow without end. Hence, we must specify an end point for the
simulations. 

`simulate_chains()` provides the `stat_max` argument for this purpose.
Above `stat_max`, the simulation is cut off. If this value is 
not specified, it assumes a value of infinity. Here, we will
assume a maximum chain size of $1000$.

```{r stat_max_setup}
#' Maximum chain size allowed
stat_max <- 1000
```

## Modelling assumptions

This exercise makes the following simplifying assumptions:

1. All cases are observed.
1. Cases are observed exactly at the time of infection.
1. There is no reporting delay.
1. Reporting rate is constant through the course of the epidemic.
1. No interventions have been implemented.
1. Population is homogeneous and well-mixed.

To summarise the whole set up so far, we are going to simulate 
each chain `r sim_rep` times, projecting cases over
`r projection_window` days after the first `r max(t0)` days, and 
assuming that no outbreak size exceeds `r stat_max` cases.

## Running the simulations

We will use the function `lapply()` to run the simulations and bind them
by rows with `dplyr::bind_rows()`.
```{r run_simulations, message=FALSE}
set.seed(1234)
sim_chain_sizes <- lapply(
  seq_len(sim_rep),
  function(sim) {
    simulate_chains(
      n_chains = length(t0),
      offspring_dist = rnbinom,
      mu = mu,
      size = size,
      statistic = "size",
      stat_max = stat_max,
      generation_time = generation_time,
      t0 = t0,
      tf = tf
    ) %>%
      mutate(sim = sim)
  }
)

sim_output <- bind_rows(sim_chain_sizes)
```

Let us view the first few rows of the simulation results.
```{r view_sim_output}
head(sim_output)
```

## Post-processing

Now, we will summarise the simulation results. 

We want to plot the individual simulated daily time series and show 
the median cases per day aggregated over all simulations.

First, we will create the daily time series per simulation by
aggregating the number of cases per day of each simulation.
```{r post_process_output}
# Daily number of cases for each simulation
incidence_ts <- sim_output %>%
  mutate(day = ceiling(time)) %>%
  count(sim, day, name = "cases") %>%
  as_tibble()

head(incidence_ts)
```

Next, we will add a date column to the results of each simulation 
set. We will use the date of the first case in the observed data 
as the reference start date.
```{r add_dates}
# Get start date from the observed data
index_date <- min(seed_cases$date)
index_date

# Add a dates column to each simulation result
incidence_ts_by_date <- incidence_ts %>%
  group_by(sim) %>%
  mutate(date = index_date + days(seq(0, n() - 1))) %>%
  ungroup()

head(incidence_ts_by_date)
```

Now we will aggregate the simulations by day and evaluate the median 
daily cases across all simulations.
```{r aggregate_simulations}
# Median daily number of cases aggregated across all simulations
median_daily_cases <- incidence_ts_by_date %>%
  group_by(date) %>%
  summarise(median_cases = median(cases)) %>%
  ungroup() %>%
  arrange(date)

head(median_daily_cases)
```

## Visualization

We will now plot the individual simulation results alongside the median
of the aggregated results.
```{r viz, fig.cap ="COVID-19 incidence in South Africa projected over a two week window in 2020. The light gray lines represent the individual simulations, the red line represents the median daily cases across all simulations, the black connected dots represent the observed data, and the dashed vertical line marks the beginning of the projection.", fig.width=6.0, fig.height=6}

# since all simulations may end at a different date, we will find the minimum
# final date for all simulations for the purposes of visualisation.
final_date <- incidence_ts_by_date %>%
  group_by(sim) %>%
  summarise(final_date = max(date), .groups = "drop") %>%
  summarise(min_final_date = min(final_date)) %>%
  pull(min_final_date)

incidence_ts_by_date <- incidence_ts_by_date %>%
  filter(date <= final_date)

median_daily_cases <- median_daily_cases %>%
  filter(date <= final_date)

ggplot(data = incidence_ts_by_date) +
  geom_line(
    aes(
      x = date,
      y = cases,
      group = sim
    ),
    color = "grey",
    linewidth = 0.2,
    alpha = 0.25
  ) +
  geom_line(
    data = median_daily_cases,
    aes(
      x = date,
      y = median_cases
    ),
    color = "tomato3",
    linewidth = 1.8
  ) +
  geom_point(
    data = covid19_sa,
    aes(
      x = date,
      y = cases
    ),
    color = "black",
    size = 1.75,
    shape = 21
  ) +
  geom_line(
    data = covid19_sa,
    aes(
      x = date,
      y = cases
    ),
    color = "black",
    linewidth = 1
  ) +
  scale_x_continuous(
    breaks = seq(
      min(incidence_ts_by_date$date),
      max(incidence_ts_by_date$date),
      5
    ),
    labels = seq(
      min(incidence_ts_by_date$date),
      max(incidence_ts_by_date$date),
      5
    )
  ) +
  scale_y_continuous(
    breaks = seq(
      0,
      max(incidence_ts_by_date$cases),
      30
    ),
    labels = seq(
      0,
      max(incidence_ts_by_date$cases),
      30
    )
  ) +
  geom_vline(
    mapping = aes(xintercept = max(seed_cases$date)),
    linetype = "dashed"
  ) +
  labs(x = "Date", y = "Daily cases")
```
## References
